---
title: "cfmDNA study: samples from group 1 and 2: diversity analysis. (Genome Biology revision 1)"
author: "Enrique Zozaya-Valdés"
date: "01/05/2021"
output:
  html_document:
    toc: yes
    number_sections: yes
    toc_float: yes
  html_notebook:
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: "hide"
urlcolor: blue
editor_options: 
  chunk_output_type: console
---

```{r PackageLoad, tidy=TRUE, message=FALSE, warning=FALSE, echo=FALSE}
library("tidyverse")
library("reshape2")
library("scales")
library("metagenomeSeq")
library("metagMisc")
library("knitr")
library("irr")
library("Biostrings")
library("ggpubr")
library("vegan")
library("ape")
library("phyloseq")
library("geepack")
library(compositions)
library(factoextra)
```

```{r FunctionsDec, message=FALSE, include=FALSE}
# Function to remove OTUs with an abundance across all samples below a given threshold. This function was taken from: http://mixomics.org/mixmc/pre-processing/.
# The function was modified to process a OTU (rows) x sample (columns) table.
low.count.removal = function(
  data, # OTU count data frame of size p (OTU) x n (sample); (rows x columns)
  percent=0.01 # cutoff chosen
){
  OTU_percent_abund = rowSums(data)*100/(sum(rowSums(data)))
  keep.otu = which(OTU_percent_abund > percent)
  data.filter = data[keep.otu,]
  return(list(data.filter = data.filter, OTU_percent_abund = OTU_percent_abund[keep.otu]))
}

## Function that changes QIIME2's Silva and Greengenes taxonomy output format to an RDP-like-fotmat
format_silva_tax <- function(taxonomy_in){
  taxonomy_out <- subset(taxonomy_in, select = Taxon)
  taxonomy_out$Taxon <- str_replace_all(taxonomy_out$Taxon, "D_\\d+__", "")
  taxonomy_out$Taxon <- str_replace_all(taxonomy_out$Taxon, "[\\w|[[:blank:]]|\\.]*(uncultured|unidentified)[\\w|[[:blank:]]|\\.]*", "")
  taxonomy_out$Taxon <- str_replace_all(taxonomy_out$Taxon, ";Incertae Sedis;", ";;")
  taxonomy_out <- separate(taxonomy_out, Taxon, c("Domain","Phylum","Class","Order","Family","Genus", "Species"), sep = ";")
  taxonomy_out[taxonomy_out == ""] <- NA
  missing_species <- which(is.na(taxonomy_out$Species))

  for(i in seq_along(missing_species)){
    deepest_assigned_pos <- sum(!is.na(taxonomy_out[missing_species[i],]))
    deepest_assigned <- taxonomy_out[missing_species[i],deepest_assigned_pos]
    missing_tax_lev <- is.na(taxonomy_out[missing_species[i],])
    if(grepl("Unassigned", deepest_assigned)){
      taxonomy_out[missing_species[i],missing_tax_lev] <- deepest_assigned
    }else{
      taxonomy_out[missing_species[i],missing_tax_lev] <- paste0(deepest_assigned, "_unclassified")
    }
  }
  taxonomy_out
}

## Function to perform PERMANOVA analysis with repeat measure-aware permutations
PERMANOVA_repeat_measures <- function(
  D, permute_within, blocks = NULL, block_data, permutations=999,
  metadata_order = c(names(permute_within), names(block_data)),
  na.rm=F) {

  # Make sure D is a dist object
  if (class(D) != "dist") {
    stop("D must be a dist object")
  }

  # Default to free permutations if blocks is not given
  if (!missing(block_data) && is.null(blocks)) {
    stop("blocks must be given if block_data is present")
  } else if (is.null(blocks)) {
    blocks <- rep(1, nrow(permute_within))
    block_data <- as.data.frame(matrix(0, nrow=1, ncol=0))
  } else if (length(unique(blocks)) == 1) {
    warning("blocks only contains one unique value")
  }

  # Ensure no metadata overlap between permute_within and block_data
  if (length(intersect(names(permute_within), names(block_data))) > 0) {
    stop("metadata is repeated across permute_within and block_data")
  }

  # Ensure that metadata_order only contains stuff in permute_within and block_data
  if(length(setdiff(metadata_order, union(names(permute_within), names(block_data)))) > 0) {
    stop("metadata_order contains metadata not in permute_within and block_data")
  }

  # Ensure that the data in permute_within matches that in dist
  ord <- rownames(as.matrix(D))
  if (length(ord) != nrow(permute_within) || length(blocks) != length(ord)) {
    stop("blocks, permute_within, and D are not the same size")
  }
  if (is.null(rownames(permute_within))) {
    warning("permute_within has no rownames - can't verify sample orders")
  } else if (!all(ord == rownames(permute_within))) {
    stop("rownames do not match between permute_within and D")
  }

  # Ensure matching between blocks and block_data
  if (any(is.na(blocks))) {
    stop("NAs are not allowed in blocks")
  }
  if (is.factor(blocks)) {
    if (length(levels(blocks)) != nrow(block_data)) {
      stop("block_data does not have as many rows as blocks has levels")
    }
    if (!is.null(rownames(block_data)) && any(rownames(block_data) != levels(blocks))) {
      stop("block_data rownames does not match the levels of blocks")
    }
    # Discard level information
    blocks <- as.numeric(blocks)
  } else if (is.numeric(blocks)) {
    if (blocks < 1 || max(blocks) > nrow(block_data)) {
      stop("Numeric blocks has indices out of range")
    }
  } else if (is.character(blocks)) {
    if (is.null(rownames(block_data)) || !all(blocks %in% rownames(block_data))) {
      stop("blocks does not match the rownames of block_data")
    }
    # Transform to numeric
    blocks <- match(blocks, rownames(block_data))
  } else {
    stop("blocks must be a numeric, factor, or character vector")
  }

  # Error out on NA metadata rather than allowing adonis to error out with
  # a totally nonsensical error message
  if (any(is.na(permute_within)) || any(is.na(block_data))) {
    if (na.rm) {
      n_prerm <- length(blocks)

      # Remove NAs in block_data
      hasna <- (rowSums(is.na(block_data)) > 0) | (sapply(split(rowSums(is.na(permute_within)) > 0, blocks), mean) == 1)
      block_data <- block_data[!hasna,, drop=F]
      keep <- !hasna[blocks]
      blocks <- cumsum(!hasna)[blocks]

      blocks <- blocks[keep]
      permute_within <- permute_within[keep,, drop=F]
      D <- as.matrix(D)[keep, keep]
      # block_data is not subset, as the rows with NAs are no longer referenced in blocks

      # Remove NAs in permute_within
      keep <- rowSums(is.na(permute_within)) == 0
      blocks <- blocks[keep]
      permute_within <- permute_within[keep,, drop=F]
      D <- as.dist(D[keep, keep])

      if (length(blocks) < ncol(permute_within) + ncol(block_data)) {
        stop(sprintf("After omitting samples with NAs, the number of samples (%d) is less than the number of metadata (%d)",
                     length(blocks), ncol(permute_within) + ncol(block_data)))
      } else if (length(blocks) < n_prerm * 0.5) {
        warning(sprintf("Removed %d samples with NA metadata", n_prerm - length(blocks)))
      }
    } else {
      stop("Some metadata is NA! adonis does not support any NA in the metadata")
    }
  }

  # Warn on some suspicious input
  persample <- apply(permute_within, 1, function(x)is.factor(x) && !any(duplicated(x)))
  if (any(persample)) {
    warning(sprintf("%s in permute_within has one DOF per sample.", colnames(permute_within)[which(persample)[1]]))
  }
  if (length(unique(blocks)) < nrow(block_data)) {
    warning("Not all blocks have a sample associated with them. Block permutations will still be performed over the full set of blocks - if this is not desired, subset block_data to only the blocks which appear in the data.")
  }
  if (!any(duplicated(blocks))) {
    warning("blocks contains no duplicated elements")
  }

  library(vegan)
  library(permute)

  # Test statistic from non-permuted data
  mtdat <- cbind(permute_within, block_data[blocks,,drop=F])
  ad <- adonis(D ~ ., permutations=0, data=mtdat[, metadata_order, drop=F])
  R2 <- ad$aov.tab$R2
  names(R2) <- rownames(ad$aov.tab)

  # Permutations
  nullsamples <- matrix(NA, nrow=length(R2), ncol=permutations)
  for (i in seq_len(permutations)) {
    within.i <- shuffle(nrow(permute_within), control=how(blocks=blocks))
    block.i <- sample(seq_len(nrow(block_data)))
    mtdat <- cbind(
      permute_within[within.i,,drop=F],
      block_data[block.i,,drop=F][blocks,,drop=F])
    perm.ad <- adonis(D ~ ., permutations=0, data=mtdat[, metadata_order, drop=F])

    nullsamples[,i] <- perm.ad$aov.tab$R2
  }

  # For residuals, test the other direction (i.e. p-value of all covariates)
  n <- length(R2)
  R2[n-1] <- 1 - R2[n-1]
  nullsamples[n-1,] <- 1 - nullsamples[n-1,]

  # P value calculation similar to adonis's
  exceedances <- rowSums(nullsamples > R2)
  P <- (exceedances + 1) / (permutations + 1)

  P[n] <- NA    # No p-values for "Total"
  ad$aov.tab$`Pr(>F)` <- P

  return (ad)
}

# Alexandra's color palette
ARS_PerCol20 = c("#FF0000",  "#FF7200", "#FFAA00", "#FFDD00", "#72d813", "#154f0d", "#06993E", "#06D8C3", "#06B2D8", "#004ECC", "#0300cc", "#6200CC", "#8E00CC", "#C500CC", "#CC0073", "#CC002C", "#BA8857", "#A04620", "#F47A00", "#381C00")
ARS_PerCol35 = c("#FF0000",  "#FF7200", "#FFAA00", "#FFDD00", "#72d813", "#154f0d", "#06993E", "#06D8C3", "#06B2D8", "#004ECC", "#0300cc", "#6200CC", "#8E00CC", "#C500CC", "#CC0073", "#CC002C", "#BA8857", "#A04620", "#F47A00", "#771155","#AA4488", "#CC99BB", "#114477", "#4477AA", "#77AADD", "#381C00", "#781156","#A51876","#D21E96","#E43FAD", "#117845","#18A55E","#1ED278","#3FE491","#6CEAAB")
```

#Data pre-processing

```{r Import_and_basic_processing, echo=FALSE, message=FALSE, warning=FALSE, results='asis', tidy=TRUE}
setwd("~/Documents/work/bioinformatics/experiments/16S_rRNA_gene/GB_revision1/R scripts for Github/Group 1 and 2/div_analysis")
data_path <- "~/Documents/work/bioinformatics/experiments/16S_rRNA_gene/GB_revision1/R scripts for Github/Group 1 and 2/data/"

load(paste0(data_path, "Group_1_and_2.data.RData"))

#Alternative: Data import from individual data files
# # Import biom file, phylogenetic tree and ASV sequences from QIIME2 
# physeq2 <- import_biom(paste0(data_path, "Group_1_and_2.biom"), parseFunction = parse_taxonomy_default, refseqfilename = paste0(data_path, "Group_1_and_2.rep_seqs.align.fasta"), treefilename = paste0(data_path, "Group_1_and_2.rep_seqs.tre"))
# # Read taxonomy file
# taxonomy <- read.table(paste0(data_path, "Group_1_and_2.rep_seqs.silva_119_99_515_806.txt"), sep = "\t", row.names = 1, header = T)
# # Create taxonomy phyloseq object
# taxonomy_phy <- tax_table(as.matrix(taxonomy))
# #Read metadata file
# metadata <- read.table(paste0(data_path, "Group_1_and_2.metadata.txt"), sep = "\t", row.names = 1, header=TRUE)
# # Create metadata phyloseq object
# metadata_phy <- sample_data(metadata)
# # Combine OTU table and taxonomy into a single phyloseq object
# physeq2 <- merge_phyloseq(physeq2, taxonomy_phy, metadata_phy)

#Remove Mock samples
physeq2 <- subset_samples(physeq, sample_type != "Mock")
physeq2 <- prune_taxa(taxa_sums(physeq2) > 0, physeq2)

##### Removal of ASVs by taxonomy  ##### 
cat('**Removal of ASVs that based on their taxonomic classification, do not form part of the microbial community:**\n\n')

cat('Number of ASVs classified as Eukaryota at the Domain level: ')
cat(sum(tax_table(physeq2)[,"Domain"] == "Eukaryota"), '\n\n')

cat('Number of ASVs classified as mitochondria at the Family level: ')
cat(sum(tax_table(physeq2)[,"Family"] == "mitochondria"), '\n\n')

cat('Number of ASVs classified as Chloroplast at the Family level: ')
cat(sum(tax_table(physeq2)[,"Class"] == "Chloroplast"), '\n\n')

# Logical vector that indicates the ASVs which based on their taxonomic classification, we want to keep.
# keep_taxa <- as.logical(!(tax_table(physeq2)[,"Domain"] == "Eukaryota" | tax_table(physeq2)[,"Domain"] == "Unassigned" | tax_table(physeq2)[,"Family"] == "mitochondria" | tax_table(physeq2)[,"Class"] == "Chloroplast"))

##### TEST: code from "manus_div_analysis.v2.GB1.Rmd" 
keep_taxa <- as.logical(!(tax_table(physeq2)[,"Domain"] == "Eukaryota" | tax_table(physeq2)[,"Family"] == "mitochondria" | tax_table(physeq2)[,"Class"] == "Chloroplast"))

cat("Total number of ASVs to keep: ")
cat(sum(keep_taxa), '\n\n') # Number of taxa to keep
cat('***\n\n')
# Removal of unwanted OTUs
physeq2 <- prune_taxa(keep_taxa, physeq2)

# Merge PCR replicates together (merge by biological sample)
physeq2_bs <- merge_samples(physeq2, "biological_sample", fun = sum)
# Because merge_samples swaps the rows and columns of the otu table, transpose it back
otu_table(physeq2_bs) <- t(otu_table(physeq2_bs))
# Re-assigning the refseq object
physeq2_bs <- merge_phyloseq(physeq2_bs, refseq(physeq2))
# Generate a new metadata table that matches the newly created merged OTU table
metadata_phy_bs <- sample_data(prune_samples(!duplicated(get_variable(physeq2, "biological_sample")), physeq2))
sample_names(metadata_phy_bs) <- metadata_phy_bs$biological_sample
sample_data(physeq2_bs) <- metadata_phy_bs

###### Filtering of OTU-tables based on OTU abundances #######

#Because the microbiome composition can vary widely depending of the sample-type, the abundance-based filter was applied to each sample-type separately. Then the orginal OTU-table was subsetted to the set of OTUs that passed the filter for at least one sample-type.

cat('**Filter of ASVs with an abundance across samples < 0.01% for each sample-type**\n\n')

# Subset OTU table to plasma
cat('Abundace-based filter of plasma samples:\n\n')
physeq_bs_pl <- subset_samples(physeq2_bs, sample_type == "Plasma")
physeq_bs_pl <- prune_taxa(taxa_sums(physeq_bs_pl) > 0, physeq_bs_pl)
# Create new phyloseq OTU object by discarding OTUs with an abundance across all samples below 0.01%
cat('* ASV abundance threshold in absolute numbers: ', sum(taxa_sums(physeq_bs_pl))*0.0001, '\n')
physeq_bs_pl_sor <- physeq_bs_pl  # sor stands for "small OTU removal"
df = low.count.removal(otu_table(physeq_bs_pl), percent=0.01)
otu_table(physeq_bs_pl_sor) <- otu_table(df$data.filter)
cat("* Number of ASVs that passed the filter: ", ntaxa(physeq_bs_pl_sor), '\n')
cat("* Percentage of ASVs that passed the filter: ", (ntaxa(physeq_bs_pl_sor) / ntaxa(physeq_bs_pl)) * 100, '\n\n')

# Subset OTU table to plasma-blank
cat('Abundace-based filter of plasma-blank samples:\n\n')
physeq_bs_plbl <- subset_samples(physeq2_bs, sample_type == "Plasma_ext_blank")
physeq_bs_plbl <- prune_taxa(taxa_sums(physeq_bs_plbl) > 0, physeq_bs_plbl)
# Create new phyloseq OTU object by discarding OTUs with an abundance across all samples below 0.01%
cat("* ASV abundance threshold in absolute numbers:", sum(taxa_sums(physeq_bs_plbl))*0.0001, '\n')
physeq_bs_plbl_sor <- physeq_bs_plbl  # sor stands for "small OTU removal"
df = low.count.removal(otu_table(physeq_bs_plbl), percent=0.01)
otu_table(physeq_bs_plbl_sor) <- otu_table(df$data.filter)
cat("* Number of ASVs that passed the filter: ", ntaxa(physeq_bs_plbl_sor), '\n')
cat("* Percentage of ASVs that passed the filter: ", (ntaxa(physeq_bs_plbl_sor) / ntaxa(physeq_bs_plbl)) * 100, '\n\n')

# Subset OTU table to stool
cat('Abundace-based filter of stool samples:\n\n')
physeq_bs_st <- subset_samples(physeq2_bs, sample_type == "Stool")
physeq_bs_st <- prune_taxa(taxa_sums(physeq_bs_st) > 0, physeq_bs_st)
# Create new phyloseq OTU object by discarding OTUs with an abundance across all samples below 0.01%
cat('* ASV abundance threshold in absolute numbers: ', sum(taxa_sums(physeq_bs_st))*0.0001, '\n')
physeq_bs_st_sor <- physeq_bs_st  # sor stands for "small OTU removal"
df = low.count.removal(otu_table(physeq_bs_st), percent=0.01)
otu_table(physeq_bs_st_sor) <- otu_table(df$data.filter)
cat('* Number of ASVs that passed the filter: ', ntaxa(physeq_bs_st_sor), '\n')
cat('* Percentage of ASVs that passed the filter: ', (ntaxa(physeq_bs_st_sor) / ntaxa(physeq_bs_st)) * 100, '\n\n')

# Subset OTU table to stool-blank
cat('Abundace-based filter of stool-blank samples:\n\n')
physeq_bs_stbl <- subset_samples(physeq2_bs, sample_type == "Stool_ext_blank")
physeq_bs_stbl <- prune_taxa(taxa_sums(physeq_bs_stbl) > 0, physeq_bs_stbl)
# Create new phyloseq OTU object by discarding OTUs with an abundance across all samples below 0.01%
cat('* ASV abundance threshold in absolute numbers: ', sum(taxa_sums(physeq_bs_stbl))*0.0001, '\n')
physeq_bs_stbl_sor <- physeq_bs_stbl  # sor stands for "small OTU removal"
df = low.count.removal(otu_table(physeq_bs_stbl), percent=0.01)
otu_table(physeq_bs_stbl_sor) <- otu_table(df$data.filter)
cat('* Number of ASVs that passed the filter: ', ntaxa(physeq_bs_stbl_sor), '\n')
cat('* Percentage of ASVs that passed the filter: ', (ntaxa(physeq_bs_stbl_sor) / ntaxa(physeq_bs_stbl)) * 100, '\n\n')

# Subset OTU table to saliva
cat('Abundace-based filter of saliva samples:\n\n')
physeq_bs_sl <- subset_samples(physeq2_bs, sample_type == "Saliva")
physeq_bs_sl <- prune_taxa(taxa_sums(physeq_bs_sl) > 0, physeq_bs_sl)
# Create new phyloseq OTU object by discarding OTUs with an abundance across all samples below 0.01%
cat('* ASV abundance threshold in absolute numbers: ', sum(taxa_sums(physeq_bs_sl))*0.0001, '\n')
physeq_bs_sl_sor <- physeq_bs_sl  # sor stands for "small OTU removal"
df = low.count.removal(otu_table(physeq_bs_sl), percent=0.01)
otu_table(physeq_bs_sl_sor) <- otu_table(df$data.filter)
cat('* Number of ASVs that passed the filter: ', ntaxa(physeq_bs_sl_sor), '\n')
cat('* Percentage of ASVs that passed the filter: ', (ntaxa(physeq_bs_sl_sor) / ntaxa(physeq_bs_sl)) * 100, '\n\n')

# Subset OTU table to saliva-blank
cat('Abundace-based filter of saliva-blank samples:\n\n')
physeq_bs_slbl <- subset_samples(physeq2_bs, sample_type == "Saliva_ext_blank")
physeq_bs_slbl <- prune_taxa(taxa_sums(physeq_bs_slbl) > 0, physeq_bs_slbl)
# Create new phyloseq OTU object by discarding OTUs with an abundance across all samples below 0.01%
cat('* ASV abundance threshold in absolute numbers: ', sum(taxa_sums(physeq_bs_slbl))*0.0001, '\n')
physeq_bs_slbl_sor <- physeq_bs_slbl  # sor stands for "small OTU removal"
df = low.count.removal(otu_table(physeq_bs_slbl), percent=0.01)
otu_table(physeq_bs_slbl_sor) <- otu_table(df$data.filter)
cat('* Number of ASVs that passed the filter: ', ntaxa(physeq_bs_slbl_sor), '\n')
cat('* Percentage of ASVs that passed the filter: ', (ntaxa(physeq_bs_slbl_sor) / ntaxa(physeq_bs_slbl)) * 100, '\n\n')

# Subset OTU table to NTC
cat('Abundace-based filter of NTCs:\n\n')
physeq_bs_NTC <- subset_samples(physeq2_bs, sample_type == "NTC")
physeq_bs_NTC <- prune_taxa(taxa_sums(physeq_bs_NTC) > 0, physeq_bs_NTC)
# Create new phyloseq OTU object by discarding OTUs with an abundance across all samples below 0.01%
cat('* ASV abundance threshold in absolute numbers: ', sum(taxa_sums(physeq_bs_NTC))*0.0001, '\n')
physeq_bs_NTC_sor <- physeq_bs_NTC  # sor stands for "small OTU removal"
df = low.count.removal(otu_table(physeq_bs_NTC), percent=0.01)
otu_table(physeq_bs_NTC_sor) <- otu_table(df$data.filter)
cat('* Number of ASVs that passed the filter: ', ntaxa(physeq_bs_NTC_sor), '\n')
cat('* Percentage of ASVs that passed the filter: ', (ntaxa(physeq_bs_NTC_sor) / ntaxa(physeq_bs_NTC)) * 100, '\n\n')

# Get the names of the OTUs that passed the low.count.removal filter for all the sample types (the union).
sor_filtered_OTUs <- unique(c(taxa_names(physeq_bs_pl_sor), taxa_names(physeq_bs_plbl_sor), taxa_names(physeq_bs_st_sor), taxa_names(physeq_bs_stbl_sor), taxa_names(physeq_bs_sl_sor), taxa_names(physeq_bs_slbl_sor), taxa_names(physeq_bs_NTC_sor)))
#Generate an OTU table that only contain the OTUs that passed the previous low.count.removal filters
physeq2_bs_sor <- prune_taxa(sor_filtered_OTUs, physeq2_bs)

cat('Abundance-based filter result for whole OTU-table:\n\n')
cat('* Number of ASVs that passed the filter for whole OTU-table: ', ntaxa(physeq2_bs_sor), '\n')
cat('* Percentage of ASVs that passed the filter for whole OTU-table: ', ntaxa(physeq2_bs_sor) / ntaxa(physeq2_bs) * 100, '\n\n')

# Histogram of sample sizes:
# sample_sizes <- as.data.frame(sample_sums(physeq2_bs_sor))
# names(sample_sizes) <- "num_of_reads"
# ggplot(sample_sizes, aes(x=num_of_reads)) + geom_histogram() + scale_x_continuous(breaks = pretty_breaks(n=20))
# Sort all samples by size
# as.data.frame(sort(sample_sums(physeq2_bs_sor)))
# Sort saliva-blanks samples by size
# as.data.frame(sort(sample_sums(physeq_bs_slbl_sor)))
# NORMALIZE OTUs by rarefaction: 4 NTC and 3 saliva_ext_blank are removed
physeq2_bs_sor_rf <- rarefy_even_depth(physeq2_bs_sor, rngseed = 711, replace=FALSE, trimOTUs = TRUE, sample.size = 128)

###### OTU TABLE SUBSETS ######

# OTU-TABLE SUBSET: plasma, blank and NTC
physeq2_bs_sor_pl_bl_NTC <- subset_samples(physeq2_bs_sor, sample_type == "Plasma" | sample_type == "Plasma_ext_blank" | sample_type == "NTC")
physeq2_bs_sor_pl_bl_NTC <- prune_taxa(taxa_sums(physeq2_bs_sor_pl_bl_NTC) > 0, physeq2_bs_sor_pl_bl_NTC)
# Sort samples by total read number:
# as.data.frame(sort(sample_sums(physeq2_bs_sor_pl_bl_NTC)))
# NORMALIZE OTUs by rarefaction: picked the min sample size that corresponds to NTC_plate4_1. This allowed to keep 11 NTC samples (6 Nov18-NTCs and 5 Jul19-NTCs) without compomising too much the rest of the samples. 9 Nov18-NTCs and 3 Jul19-NTCs were removed.
physeq2_bs_sor_pl_bl_NTC_rf <- rarefy_even_depth(physeq2_bs_sor_pl_bl_NTC, rngseed = 711, replace=FALSE, trimOTUs = TRUE, sample.size = 5306)

# OTU-TABLE SUBSET: plasma, blank
physeq2_bs_sor_pl_bl <- subset_samples(physeq2_bs_sor, sample_type == "Plasma" | sample_type == "Plasma_ext_blank")
physeq2_bs_sor_pl_bl <- prune_taxa(taxa_sums(physeq2_bs_sor_pl_bl) > 0, physeq2_bs_sor_pl_bl)
# Sort samples by total read number:
# as.data.frame(sort(sample_sums(physeq2_bs_sor_pl_bl)))
# NORMALIZE OTUs by rarefaction to min. sample size
physeq2_bs_sor_pl_bl_rf <- rarefy_even_depth(physeq2_bs_sor_pl_bl, rngseed = 711, replace=FALSE, trimOTUs = TRUE)

# OTU-TABLE SUBSET: all samples excepting Mock, Saliva and Saliva-blanks
physeq2_bs_sor_subset2 <- subset_samples(physeq2_bs_sor, sample_type != "Saliva" & sample_type != "Saliva_ext_blank")
physeq2_bs_sor_subset2 <- prune_taxa(taxa_sums(physeq2_bs_sor_subset2) > 0, physeq2_bs_sor_subset2)
# Sort samples by total read number:
# as.data.frame(sort(sample_sums(physeq2_bs_sor_subset2)))
# NORMALIZE OTUs by rarefaction: picked the min sample size that corresponds to NTC_plate5_3 because this was the smallest sample with more than 1000 reads. This removes 6 Stool_ext_blanks and 7 NTCs
physeq2_bs_sor_subset2_rf <- rarefy_even_depth(physeq2_bs_sor_subset2, rngseed = 711, replace=FALSE, trimOTUs = TRUE, sample.size = 1309)
# OTU-TABLE SUBSET: all samples excepting Mock and blanks
physeq2_bs_sor_subset3 <- subset_samples(physeq2_bs_sor, sample_type != "Saliva_ext_blank" & sample_type != "Plasma_ext_blank" & sample_type != "Stool_ext_blank" & sample_type != "NTC")
physeq2_bs_sor_subset3 <- prune_taxa(taxa_sums(physeq2_bs_sor_subset3) > 0, physeq2_bs_sor_subset3)
# Sort samples by total read number:
# as.data.frame(sort(sample_sums(physeq2_bs_sor_subset3)))
# NORMALIZE OTUs by rarefaction
physeq2_bs_sor_subset3_rf <- rarefy_even_depth(physeq2_bs_sor_subset3, rngseed = 711, replace=FALSE, trimOTUs = TRUE)
```

#Sample sizes assessment

```{r sample_sizes, echo=FALSE, results="hide", message=FALSE, fig.show="show",  warning=FALSE}
system("mkdir sample_sizes")

sample_sizes_w_metadata <- cbind(sample_data(physeq2_bs_sor), sample_size = sample_sums(physeq2_bs_sor))
y <- dplyr::group_by(sample_sizes_w_metadata, sample_type)
cat('Average number of reads per sample-type\n\n')
dplyr::summarize(y, ave = mean(sample_size))

## Plot with ggpubr: sample size (number of reads) per sample type: Supp. figure
ggpubr_bx <- ggboxplot(sample_sizes_w_metadata, x="sample_type", y="sample_size", color="sample_type", 
                       add ="mean",  size=1, width=0.5, 
                       order=c("NTC","Saliva","Saliva_ext_blank","Stool","Stool_ext_blank","Plasma","Plasma_ext_blank"), 
                       palette = c("light blue","#329E26","#7CCE73","#9E5524","#BC8F71","red", "#F78F8F"), xlab="sample type", 
                       ylab="number of reads", legend="none", ggtheme = theme_bw()) 

ggpubr_bx + scale_x_discrete(labels=c("NTC","saliva","saliva-DENC","stool","stool-DENC","plasma","plasma-DENC"))  + scale_y_continuous(breaks = pretty_breaks(n=6), labels=comma_format())  + theme(panel.grid.major.x = element_blank(), axis.text =element_text(size=11), axis.title = element_text(size=14), plot.title=element_text(size=10, face="bold"))  + ggtitle("Figure S2. Number of 16S rRNA gene quality-filtered reads obtained across sample types") 
ggsave("./sample_sizes/Figure S2.pdf", width = 7, height = 4.5)
```

#Mock communities

```{r Mock, echo=FALSE, results="hide", message=FALSE, fig.show="show",  warning=FALSE}
system("mkdir mock")

# Get a subset of the orginal OTU table that only contains the Mock samples
physeq_Mock <- subset_samples(physeq, sample_type == "Mock")
physeq_Mock <- prune_taxa(taxa_sums(physeq_Mock) > 0, physeq_Mock)

# Create new phyloseq OTU object by discarding OTUs with an abundance across all samples below 0.01%
physeq_Mock_sor <- physeq_Mock
df = low.count.removal(otu_table(physeq_Mock), percent=0.01)
otu_table(physeq_Mock_sor) <- otu_table(df$data.filter)

# Rarified OTU table to the minimum sample size 
#sort(sample_sums(physeq_Mock_sor)) #Get the sample sizes and sort them
physeq_Mock_sor_rf <- rarefy_even_depth(physeq_Mock_sor, rngseed = 711, replace=FALSE, trimOTUs = TRUE, sample.size = 9593)
#estimate_richness(physeq_Mock_sor_rf, measures = c("Observed"))

# Rank-abundance plot
total_taxa_abund <- sort(taxa_sums(physeq_Mock_sor_rf), TRUE, index.return=T)
genera_sorted <- as.character(tax_table(physeq_Mock_sor_rf)[total_taxa_abund$ix,"Genus"])
total_taxa_abund.df <- data.frame(OTU = names(total_taxa_abund$x), abundance = total_taxa_abund$x/nsamples(physeq_Mock_sor_rf))
mock_rank_abund_plot <- ggplot(total_taxa_abund.df, aes(x = OTU, y = abundance)) + geom_bar(stat="identity") + scale_x_discrete(limits=names(total_taxa_abund$x), labels = genera_sorted) + theme(axis.text =element_text(size=10), axis.text.x = element_text(angle=45, hjust=1, vjust=1), plot.margin = unit(c(0.5,0.5,1,2.5),"cm"), axis.title = element_text(size=12),  plot.title=element_text(size=10, face="bold")) + xlab("ASV") + ggtitle("Figure S3 A. Taxonomic profile of a 20 strain evenly mixed\n mock community sequenced alongside study samples")
mock_rank_abund_plot
ggsave("./mock/Figure S3 A.pdf", width = 10, height = 6)

# Bar plot of OTUs of Mock samples colored by genus
physeq_Mock_sor_rf_top20 <- prune_taxa(names(total_taxa_abund$x[1:20]), physeq_Mock_sor_rf)
plot_bar(physeq_Mock_sor_rf_top20, fill="Genus") + theme(axis.text.x = element_blank(), axis.text.y = element_text(size=12), axis.title = element_text(size=10, face="bold"), legend.text = element_text(size=10), plot.title=element_text(size=10, face="bold")) + scale_colour_manual(values=ARS_PerCol35) + scale_fill_manual(values=ARS_PerCol35) + ggtitle("Figure S3 B. Taxonomic profile of a 20 strain evenly mixed\n mock community sequenced alongside study samples")
ggsave("./mock/Figure S3 B.pdf", width = 10, height = 5)
```

#Alpha diversity

For ploting alpha-diversity measurments that included NTCs, the OTU-table was rarefied at a minimum sample size of 5,306 reads. This sample size allowed to keep 11 NTCs.

```{r Alpha_div, echo=FALSE, results="hide", fig.show="show"}
system("mkdir alpha_div")

# ALPHA-DIVERISITY OF PLASMA, BLANKS AND NTC by "DNA ext batch" and "sample type"
alpha_div_plot <- plot_richness(physeq2_bs_sor_pl_bl_NTC_rf, x = "plasma_ext_batch", measures=c("Observed", "InvSimpson"), color="sample_type")
# Remove the geom_point layer from the ggplot object
alpha_div_plot$layers <- alpha_div_plot$layers[-1]
alpha_div_plot + geom_boxplot(data=alpha_div_plot$data, aes(color=sample_type), alpha=0.1, size=1) + ggtitle("Figure 3D. Community structure of cell free microbial DNA\n in plasma and patient-matching stool and saliva") + theme(axis.text = element_text(size=14), axis.text.x = element_text(angle = 0, hjust = 0.5), axis.title = element_text(size=14), plot.title = element_text(size=12), strip.text = element_text(face="bold", size=rel(1.2)), legend.title = element_text(size=14, face="bold"), legend.text = element_text(size=14)) + labs(color="sample type") + scale_color_manual(values=c("light blue","red", "#F78F8F"), labels = c("NTC","plasma","plasma-DENC")) + scale_x_discrete(breaks=c(c("A","B","C","D","E"), NA), labels = c("A","B","C","D","E","NTC")) + xlab("Extraction batch/NTC")
ggsave("./alpha_div/Figure 3D.pdf", width = 8, height = 5)
```

Hypothesis test of alpha-diversity measurments based on generalized estimating equations (GEE). This allowed to control for repeated-measurments as in the patient-matching plasma samples of the DEBs A and B.

```{r echo=FALSE, warning=FALSE}
# Get alpha diversity estimates for every sample
pl_bl_alpha_div <- estimate_richness(physeq2_bs_sor_pl_bl_NTC_rf, measures = c("Observed","Simpson"))
pl_bl_alpha_div <- cbind(sample_data(physeq2_bs_sor_pl_bl_NTC_rf), pl_bl_alpha_div)
pl_bl_alpha_div <- subset(pl_bl_alpha_div, sample_type %in% c("Plasma","Plasma_ext_blank"))
pl_bl_alpha_div$sample_type <- droplevels(pl_bl_alpha_div$sample_type)

### GEE test of interaction between sample_type and plasma_ext_batch ###

data_test <- pl_bl_alpha_div
#Assign the sample_id as biomarker_ID for blanks
data_test$biomarker_ID <- as.character(data_test$biomarker_ID)
data_test$biomarker_ID[data_test$sample_type == "Plasma_ext_blank"] <- as.character(data_test$biological_sample[data_test$sample_type == "Plasma_ext_blank"])
data_test$biomarker_ID <- as.factor(data_test$biomarker_ID)

#Number of OTUs by sample_type test across entire cohort 
cat("Test of the difference between the richness of plasma vs DENCs across DEBs and of the interaction between sample-type and DEB\n\n")
geeglm_test <- geeglm(Observed ~ plasma_ext_batch * sample_type, data = data_test, id=biomarker_ID, corstr="exchangeable", na.action=na.omit)
anova(geeglm_test)
#InvSimpson by sample_type test across entire cohort 
cat("Test of the difference between the Simpson index of plasma vs DENCs across DEBs and of the interaction between sample-type and DEB\n\n")
geeglm_test <- geeglm(Simpson ~ plasma_ext_batch * sample_type, data = data_test, id=biomarker_ID, corstr="exchangeable", na.action=na.omit)
anova(geeglm_test)

#Number of OTUs, Simpson and Shannon tests per DEB
data_test.s <- split(data_test, data_test$plasma_ext_batch)
geeglm.test.out <- sapply(data_test.s, function(x){
  geeglm.test.out.all <- numeric()
  geeglm.test.out <- geeglm(Observed ~ sample_type, data = x, id=biomarker_ID, corstr="exchangeable", na.action=na.omit)
  geeglm.test.out.all[1] <- summary(geeglm.test.out)$coefficients[2,4]
  geeglm.test.out <- geeglm(Simpson ~ sample_type, data = x, id=biomarker_ID, corstr="exchangeable", na.action=na.omit)
  geeglm.test.out.all[2] <- summary(geeglm.test.out)$coefficients[2,4]
  names(geeglm.test.out.all) <- c("Observed","Simpson")
  geeglm.test.out.all
})

kable(geeglm.test.out, caption = "Geeglm test p-values for the comparisons of Plasma vs Plasma-blanks within each DNA extraction batch")
```

#Beta diversity

##All sample types

Non-metric Multidimensional Scaling (nMDS) of pairwise Bray-Curtis dissimilarities calculated from microbial community profiles for patient-matching plasma (batch A and B only), stool and saliva samples and respective DENCs at the ASV-level. For this analysis the OTU-table was rarefied and square-root transformed.

```{r Beta_div_all_sample_types, echo=FALSE, results="hide", message=FALSE, fig.show="show", warning=FALSE}
system("mkdir beta_div")
# Set a publication ggplot theme for ordination plots
pub_theme <- theme(axis.title = element_text(size=14), axis.text = element_text(size=14), plot.title = element_text(size=10), legend.title = element_text(size=14, face="bold"), legend.text = element_text(size=14))

#Subset OTU table to Nov-18 samples
physeq2_bs_sor_Nov18_rf <- subset_samples(physeq2_bs_sor_rf, seq_run == "Nov_2018")
physeq2_bs_sor_Nov18_rf <- prune_taxa(taxa_sums(physeq2_bs_sor_Nov18_rf) > 0, physeq2_bs_sor_Nov18_rf)
#### Data transformations
physeq2_bs_sor_Nov18_rf_sqrt <- transform_sample_counts(physeq2_bs_sor_Nov18_rf, function(OTU) sqrt(OTU))
#NMDS plot (Bray-Curtis) by sample type: sqrt transformed
NMDS<-ordinate(physeq2_bs_sor_Nov18_rf_sqrt, method="NMDS", distance="bray")
plot_NMDS<-plot_ordination(physeq2_bs_sor_Nov18_rf_sqrt, NMDS, type="samples", color="sample_type")
plot_NMDS + geom_point(size = 2) + ggtitle(label ="Figure 3B. Community structure of cell free microbial DNA in \n plasma and patient-matching stool and saliva") + pub_theme + scale_color_manual(values=c("light blue","red", "#F78F8F","#329E26","#7CCE73","#9E5524","#BC8F71"), labels=c("NTC","plasma","plasma-DENC","saliva","saliva-DENC","stool","stool-DENC")) + labs(colour="sample type")
ggsave("./beta_div/Figure 3B.pdf", width = 7, height = 5)
```

**PERMANOVA test of the overall effect of sample type (plasma, stool and saliva) on the community structure**

Technical note: OTU table rarified to 12,049 reads was used. No data transformation was applied. Bray-Curtis dissimilarity used. 

```{r RMAP_overall, echo=FALSE, warning=FALSE, message=FALSE}
# Subset OTU table to Nov-2018 seq-run and sample types plasma, stool and saliva
physeq2_bs_sor_Nov18_no_blanks <- subset_samples(physeq2_bs_sor, seq_run == "Nov_2018" &  sample_type %in% c("Plasma","Stool","Saliva"))
physeq2_bs_sor_Nov18_no_blanks <- prune_taxa(taxa_sums(physeq2_bs_sor_Nov18_no_blanks) > 0, physeq2_bs_sor_Nov18_no_blanks)
#Rarefy to even depth
physeq2_bs_sor_Nov18_no_blanks_rf <- rarefy_even_depth(physeq2_bs_sor_Nov18_no_blanks, rngseed = 711, replace=FALSE, trimOTUs = TRUE)
#Calculate Bray-Curtis dissimilarity
BC_dist <- distance(physeq2_bs_sor_Nov18_no_blanks_rf, method="bray")
Meta_df <- data.frame(sample_data(physeq2_bs_sor_Nov18_no_blanks_rf))
```

```{r RMAP, echo=FALSE}
# Create "reapeted mesurment factor" for blocking data
Meta_df$repeat_meas <- Meta_df$biomarker_ID
#permute_within
ST <- Meta_df$sample_type
PW <- as.data.frame(ST)
rownames(PW) <- rownames(Meta_df)
#blocks
BS <- as.vector(Meta_df$repeat_meas)
#block_data
Meta_subj <- Meta_df[order(Meta_df$repeat_meas),]
Meta_subj <- Meta_subj[!duplicated(Meta_subj$repeat_meas),, drop=F]
RM <- Meta_subj$repeat_meas
Meta_sub <- as.data.frame(RM)
rownames(Meta_sub) <- Meta_subj$repeat_meas
```

```{r RMAP_results, echo=FALSE}
#Performing test
RMAPres <- PERMANOVA_repeat_measures(
    D = BC_dist, permutations=1000,
    permute_within=PW,
    blocks=BS, block_data=Meta_sub,
    metadata_order = c(names(Meta_sub), names(PW)))
RMAPres
```

**Repeated measurment aware PERMANOVA: Plasma vs Stool**

```{r RMAP_pairwise1, echo=FALSE, warning=FALSE, message=FALSE}
physeq_RMAP <- subset_samples(physeq2_bs_sor_Nov18_no_blanks_rf, sample_type == "Plasma" | sample_type == "Stool")
physeq_RMAP <- prune_taxa(taxa_sums(physeq_RMAP) > 0, physeq_RMAP)
# Normalize OTUs by rarefying at the minimum sample size.
physeq_RMAP <- rarefy_even_depth(physeq_RMAP, rngseed = 711, replace=FALSE, trimOTUs = TRUE)
BC_dist <- distance(physeq_RMAP, method="bray")
# Metadata
Meta_df <- data.frame(sample_data(physeq_RMAP))
```

```{r ref.label='RMAP', echo=FALSE}
```

```{r ref.label='RMAP_results', echo=FALSE}
```

**Repeated measurment aware PERMANOVA: Plasma vs Saliva**

```{r RMAP_pairwise2, echo=FALSE, warning=FALSE, message=FALSE}
physeq_RMAP <- subset_samples(physeq2_bs_sor_Nov18_no_blanks_rf, sample_type == "Plasma" | sample_type == "Saliva")
physeq_RMAP <- prune_taxa(taxa_sums(physeq_RMAP) > 0, physeq_RMAP)
# Normalize OTUs by rarefying at the minimum sample size.
physeq_RMAP <- rarefy_even_depth(physeq_RMAP, rngseed = 711, replace=FALSE, trimOTUs = TRUE)
BC_dist <- distance(physeq_RMAP, method="bray")
# Metadata
Meta_df <- data.frame(sample_data(physeq_RMAP))
```

```{r ref.label='RMAP', echo=FALSE}
```

```{r ref.label='RMAP_results', echo=FALSE}
```

**Repeated measurment aware PERMANOVA: Stool vs Saliva**

```{r RMAP_pairwise3, echo=FALSE, warning=FALSE, message=FALSE}
physeq_RMAP <- subset_samples(physeq2_bs_sor_Nov18_no_blanks_rf, sample_type == "Stool" | sample_type == "Saliva")
physeq_RMAP <- prune_taxa(taxa_sums(physeq_RMAP) > 0, physeq_RMAP)
# Normalize OTUs by rarefying at the minimum sample size.
physeq_RMAP <- rarefy_even_depth(physeq_RMAP, rngseed = 711, replace=FALSE, trimOTUs = TRUE)
BC_dist <- distance(physeq_RMAP, method="bray")
# Metadata
Meta_df <- data.frame(sample_data(physeq_RMAP))
```

```{r ref.label='RMAP', echo=FALSE}
```

```{r ref.label='RMAP_results', echo=FALSE}
```

**PERMANOVA: Stool vs Stool_ext_blank**

```{r Adonis, echo=FALSE, warning=FALSE, message=FALSE}
# Subset OTU table
physeq2_bs_sor_st_bl <- subset_samples(physeq2_bs_sor, sample_type %in% c("Stool","Stool_ext_blank") )
physeq2_bs_sor_st_bl <- prune_taxa(taxa_sums(physeq2_bs_sor_st_bl) > 0, physeq2_bs_sor_st_bl)
# Normalize OTUs by rarefying at the minimum sample size.
physeq2_bs_sor_st_bl_rf <- rarefy_even_depth(physeq2_bs_sor_st_bl, rngseed = 711, replace=FALSE, trimOTUs = TRUE)

#Adonis test: sample_type controlled by ext_date
metadata <- data.frame(sample_data(physeq2_bs_sor_st_bl_rf))
BC_dist <- distance(physeq2_bs_sor_st_bl_rf, method="bray")
adonis.out <- adonis(BC_dist ~ ext_date * sample_type, data=metadata)
adonis.out
```

**PERMANOVA: Saliva vs Saliva_ext_blank**

```{r Adonis2, echo=FALSE, warning=FALSE, message=FALSE}
# Subset OTU table
physeq2_bs_sor_sl_bl <- subset_samples(physeq2_bs_sor, sample_type %in% c("Saliva","Saliva_ext_blank") )
physeq2_bs_sor_sl_bl <- prune_taxa(taxa_sums(physeq2_bs_sor_sl_bl) > 0, physeq2_bs_sor_sl_bl)
# Normalize OTUs by rarefying at the minimum sample size.
physeq2_bs_sor_sl_bl_rf <- rarefy_even_depth(physeq2_bs_sor_sl_bl, rngseed = 711, replace=FALSE, trimOTUs = TRUE, sample.size = 119)

#Adonis test: sample_type controlled by ext_date
metadata <- data.frame(sample_data(physeq2_bs_sor_sl_bl_rf))
BC_dist <- distance(physeq2_bs_sor_sl_bl_rf, method="bray")
adonis.out <- adonis(BC_dist ~ ext_date * sample_type, data=metadata)
adonis.out
```

##Plasma and plasma_blanks across groups 1 and 2

```{r Beta_div_plasma, echo=FALSE, results="hide", message=FALSE, fig.show="show",  warning=FALSE}
#### ALL PLASMA AND PLASMA-BLANKS

# Data transformations
physeq2_bs_sor_pl_bl_rf_sqrt <- transform_sample_counts(physeq2_bs_sor_pl_bl_rf, function(OTU) sqrt(OTU))

### Hierarchical clustering ###

#UPGMA clustering plot
BC_dist <- phyloseq::distance(physeq2_bs_sor_pl_bl_rf_sqrt, method="bray")
BC_dist_hclust <- hclust(BC_dist, method="average")
BC_dist_hclust.p <- as.phylo(BC_dist_hclust)
write.tree(BC_dist_hclust.p, file="./beta_div/hclust_dendogram_for_Fig3C_and_FigS5.newick")

#Same UPGMA clustering but using Aitchison distance
data <- as.data.frame(t(otu_table(physeq2_bs_sor_pl_bl)))
data.clr <- clr(data + 1)
data.clr.dist <- dist(data.clr, method = "euclidean")
data.clr.dist.hclust <- hclust(data.clr.dist, method="average")
data.clr.dist.hclust.p <- as.phylo(data.clr.dist.hclust)
write.tree(data.clr.dist.hclust.p, file="./beta_div/hclust_dendogram_for_FigS7.newick")
```

###Bray-Curtis based PERMANOVA

**RMA-PERMANOVA test of the effect of DNA-ext-date (controlling by sample-type; with interaction)**

```{r RMAP_overall2, echo=FALSE}
BC_dist <- distance(physeq2_bs_sor_pl_bl_rf, method="bray")
Meta_df <- data.frame(sample_data(physeq2_bs_sor_pl_bl_rf))
```

```{r RMAP2, echo=FALSE}
# Create "reapeted mesurment factor" for blocking data
Meta_df$repeat_meas <- as.character(Meta_df$biomarker_ID)
Meta_df$repeat_meas[is.na(Meta_df$repeat_meas)] <- as.character(Meta_df$biological_sample[is.na(Meta_df$repeat_meas)])
Meta_df$repeat_meas <- as.factor(Meta_df$repeat_meas)
#permute_within
ED <- Meta_df$ext_date
PW <- as.data.frame(ED)
rownames(PW) <- rownames(Meta_df)
# INTERACTION FACTOR
PW$ST_ED <- factor(paste(Meta_df$sample_type, PW$ED, sep = "."))
#blocks
BS <- as.vector(Meta_df$repeat_meas)
#block_data
Meta_subj <- Meta_df[order(Meta_df$repeat_meas),]
Meta_subj <- Meta_subj[!duplicated(Meta_subj$repeat_meas),, drop=F]
#RM <- Meta_subj$repeat_meas
#Meta_sub <- as.data.frame(RM)
ST <- Meta_subj$sample_type
Meta_sub <- as.data.frame(ST)
rownames(Meta_sub) <- Meta_subj$repeat_meas
```

```{r RMAP_results2, echo=FALSE}
#Performing test
RMAPres <- PERMANOVA_repeat_measures(
    D = BC_dist, permutations=1000,
    permute_within=PW,
    blocks=BS, block_data=Meta_sub,
    metadata_order = c(names(Meta_sub), names(PW)))
RMAPres
```

**RMA-PERMANOVA test of the difference between plasma and plasma-blanks (controlling for DNA-ext-date; without interaction)**

```{r RMAP3, echo=FALSE}
# Create "reapeted mesurment factor" for blocking data
Meta_df$repeat_meas <- as.character(Meta_df$biomarker_ID)
Meta_df$repeat_meas[is.na(Meta_df$repeat_meas)] <- as.character(Meta_df$biological_sample[is.na(Meta_df$repeat_meas)])
Meta_df$repeat_meas <- as.factor(Meta_df$repeat_meas)
#permute_within
ED <- Meta_df$ext_date
PW <- as.data.frame(ED)
rownames(PW) <- rownames(Meta_df)
#blocks
BS <- as.vector(Meta_df$repeat_meas)
#block_data
Meta_subj <- Meta_df[order(Meta_df$repeat_meas),]
Meta_subj <- Meta_subj[!duplicated(Meta_subj$repeat_meas),, drop=F]
#RM <- Meta_subj$repeat_meas
#Meta_sub <- as.data.frame(RM)
ST <- Meta_subj$sample_type
Meta_sub <- as.data.frame(ST)
rownames(Meta_sub) <- Meta_subj$repeat_meas
```

```{r RMAP_results3, echo=FALSE}
RMAPres <- PERMANOVA_repeat_measures(
    D = BC_dist, permutations=1000,
    permute_within=PW,
    blocks=BS, block_data=Meta_sub)
RMAPres
```

**RMA-PERMANOVA test of the effect of DNA-extraction-batch (controlling for sample-type; with interaction)**

```{r RMAP4, echo=FALSE}
#permute_within
DEB <- Meta_df$plasma_ext_batch
PW <- as.data.frame(DEB)
rownames(PW) <- rownames(Meta_df)
# INTERACTION FACTOR
PW$ST_DEB <- factor(paste(Meta_df$sample_type, PW$DEB, sep = "."))
#blocks
BS <- as.vector(Meta_df$repeat_meas)
#block_data
Meta_subj <- Meta_df[order(Meta_df$repeat_meas),]
Meta_subj <- Meta_subj[!duplicated(Meta_subj$repeat_meas),, drop=F]
#RM <- Meta_subj$repeat_meas
#Meta_sub <- as.data.frame(RM)
ST <- Meta_subj$sample_type
Meta_sub <- as.data.frame(ST)
rownames(Meta_sub) <- Meta_subj$repeat_meas
```

```{r RMAP_results4, echo=FALSE}
#Performing test
RMAPres <- PERMANOVA_repeat_measures(
    D = BC_dist, permutations=1000,
    permute_within=PW,
    blocks=BS, block_data=Meta_sub,
    metadata_order = c(names(Meta_sub), names(PW)))
RMAPres
```

**RMA-PERMANOVA test of the difference between plasma and plasma-blanks (controlling for DNA-extraction-batch; with interaction)**

```{r echo=FALSE}
#Performing test
RMAPres <- PERMANOVA_repeat_measures(
    D = BC_dist, permutations=1000,
    permute_within=PW,
    blocks=BS, block_data=Meta_sub,
    metadata_order = c("DEB","ST","ST_DEB"))
RMAPres
```

**RMA-PERMANOVA test of the effect of sequencing-run (controlling for sample-type; with interaction) **

```{r RMAP5, echo=FALSE}
#permute_within
SR <- Meta_df$seq_run
PW <- as.data.frame(SR)
rownames(PW) <- rownames(Meta_df)
# INTERACTION FACTOR
PW$ST_SR <- factor(paste(Meta_df$sample_type, PW$SR, sep = "."))
#blocks
BS <- as.vector(Meta_df$repeat_meas)
#block_data
Meta_subj <- Meta_df[order(Meta_df$repeat_meas),]
Meta_subj <- Meta_subj[!duplicated(Meta_subj$repeat_meas),, drop=F]
#RM <- Meta_subj$repeat_meas
#Meta_sub <- as.data.frame(RM)
ST <- Meta_subj$sample_type
Meta_sub <- as.data.frame(ST)
rownames(Meta_sub) <- Meta_subj$repeat_meas
```

```{r RMAP_results5, echo=FALSE}
#Performing test
RMAPres <- PERMANOVA_repeat_measures(
    D = BC_dist, permutations=1000,
    permute_within=PW,
    blocks=BS, block_data=Meta_sub,
    metadata_order = c(names(Meta_sub), names(PW)))
RMAPres
```

**RMA-PERMANOVA test of the difference between plasma and plasma-blanks (controlling for DNA-extraction-batch; with interaction)**

```{r echo=FALSE}
#Performing test
RMAPres <- PERMANOVA_repeat_measures(
    D = BC_dist, permutations=1000,
    permute_within=PW,
    blocks=BS, block_data=Meta_sub,
    metadata_order = c("SR","ST","ST_SR"))
RMAPres
```

###Aitchison-distance based PERMANOVA

**RMA-PERMANOVA test of the effect of DNA-ext-date (controlling by sample-type; with interaction)**

```{r echo=FALSE}
# Create "reapeted mesurment factor" for blocking data
Meta_df$repeat_meas <- as.character(Meta_df$biomarker_ID)
Meta_df$repeat_meas[is.na(Meta_df$repeat_meas)] <- as.character(Meta_df$biological_sample[is.na(Meta_df$repeat_meas)])
Meta_df$repeat_meas <- as.factor(Meta_df$repeat_meas)
#permute_within
ED <- Meta_df$ext_date
PW <- as.data.frame(ED)
rownames(PW) <- rownames(Meta_df)
# INTERACTION FACTOR
PW$ST_ED <- factor(paste(Meta_df$sample_type, PW$ED, sep = "."))
#blocks
BS <- as.vector(Meta_df$repeat_meas)
#block_data
Meta_subj <- Meta_df[order(Meta_df$repeat_meas),]
Meta_subj <- Meta_subj[!duplicated(Meta_subj$repeat_meas),, drop=F]
#RM <- Meta_subj$repeat_meas
#Meta_sub <- as.data.frame(RM)
ST <- Meta_subj$sample_type
Meta_sub <- as.data.frame(ST)
rownames(Meta_sub) <- Meta_subj$repeat_meas
```

```{r echo=FALSE}
#Performing test
RMAPres <- PERMANOVA_repeat_measures(
    D = data.clr.dist, permutations=1000,
    permute_within=PW,
    blocks=BS, block_data=Meta_sub,
    metadata_order = c(names(Meta_sub), names(PW)))
RMAPres

```

**RMA-PERMANOVA test of the difference between plasma and plasma-blanks (controlling for DNA-ext-date; without interaction)**

```{r echo=FALSE}
# Create "reapeted mesurment factor" for blocking data
Meta_df$repeat_meas <- as.character(Meta_df$biomarker_ID)
Meta_df$repeat_meas[is.na(Meta_df$repeat_meas)] <- as.character(Meta_df$biological_sample[is.na(Meta_df$repeat_meas)])
Meta_df$repeat_meas <- as.factor(Meta_df$repeat_meas)
#permute_within
ED <- Meta_df$ext_date
PW <- as.data.frame(ED)
rownames(PW) <- rownames(Meta_df)
#blocks
BS <- as.vector(Meta_df$repeat_meas)
#block_data
Meta_subj <- Meta_df[order(Meta_df$repeat_meas),]
Meta_subj <- Meta_subj[!duplicated(Meta_subj$repeat_meas),, drop=F]
#RM <- Meta_subj$repeat_meas
#Meta_sub <- as.data.frame(RM)
ST <- Meta_subj$sample_type
Meta_sub <- as.data.frame(ST)
rownames(Meta_sub) <- Meta_subj$repeat_meas
```

```{r echo=FALSE}
RMAPres <- PERMANOVA_repeat_measures(
    D = data.clr.dist, permutations=1000,
    permute_within=PW,
    blocks=BS, block_data=Meta_sub)
RMAPres
```

**RMA-PERMANOVA test of the effect of DNA-extraction-batch (controlling for sample-type; with interaction)**

```{r echo=FALSE}
#permute_within
DEB <- Meta_df$plasma_ext_batch
PW <- as.data.frame(DEB)
rownames(PW) <- rownames(Meta_df)
# INTERACTION FACTOR
PW$ST_DEB <- factor(paste(Meta_df$sample_type, PW$DEB, sep = "."))
#blocks
BS <- as.vector(Meta_df$repeat_meas)
#block_data
Meta_subj <- Meta_df[order(Meta_df$repeat_meas),]
Meta_subj <- Meta_subj[!duplicated(Meta_subj$repeat_meas),, drop=F]
#RM <- Meta_subj$repeat_meas
#Meta_sub <- as.data.frame(RM)
ST <- Meta_subj$sample_type
Meta_sub <- as.data.frame(ST)
rownames(Meta_sub) <- Meta_subj$repeat_meas
```

```{r echo=FALSE}
#Performing test
RMAPres <- PERMANOVA_repeat_measures(
    D = data.clr.dist, permutations=1000,
    permute_within=PW,
    blocks=BS, block_data=Meta_sub,
    metadata_order = c(names(Meta_sub), names(PW)))
RMAPres
```

**RMA-PERMANOVA test of the difference between plasma and plasma-blanks (controlling for DNA-extraction-batch; with interaction)**

```{r echo=FALSE}
#Performing test
RMAPres <- PERMANOVA_repeat_measures(
    D = data.clr.dist, permutations=1000,
    permute_within=PW,
    blocks=BS, block_data=Meta_sub,
    metadata_order = c("DEB","ST","ST_DEB"))
RMAPres
```

**RMA-PERMANOVA test of the effect of sequencing-run (controlling for sample-type; with interaction) **

```{r echo=FALSE}
#permute_within
SR <- Meta_df$seq_run
PW <- as.data.frame(SR)
rownames(PW) <- rownames(Meta_df)
# INTERACTION FACTOR
PW$ST_SR <- factor(paste(Meta_df$sample_type, PW$SR, sep = "."))
#blocks
BS <- as.vector(Meta_df$repeat_meas)
#block_data
Meta_subj <- Meta_df[order(Meta_df$repeat_meas),]
Meta_subj <- Meta_subj[!duplicated(Meta_subj$repeat_meas),, drop=F]
#RM <- Meta_subj$repeat_meas
#Meta_sub <- as.data.frame(RM)
ST <- Meta_subj$sample_type
Meta_sub <- as.data.frame(ST)
rownames(Meta_sub) <- Meta_subj$repeat_meas
```

```{r echo=FALSE}
#Performing test
RMAPres <- PERMANOVA_repeat_measures(
    D = data.clr.dist, permutations=1000,
    permute_within=PW,
    blocks=BS, block_data=Meta_sub,
    metadata_order = c(names(Meta_sub), names(PW)))
RMAPres
```

**RMA-PERMANOVA test of the difference between plasma and plasma-blanks (controlling for sequencing-run; with interaction)**

```{r echo=FALSE}
#Performing test
RMAPres <- PERMANOVA_repeat_measures(
    D = data.clr.dist, permutations=1000,
    permute_within=PW,
    blocks=BS, block_data=Meta_sub,
    metadata_order = c("SR","ST","ST_SR"))
RMAPres
```

#Taxonomy

##Group 1

```{r taxonomy, echo=FALSE, results="hide", message=FALSE, fig.show="show",  warning=FALSE, fig.height=9, fig.width=8}
system("mkdir taxonomy")

### NOV-18 DATA ### 

#Create new factor by combining plasma_ext_batch with sample_type
fact <- as.character(get_variable(physeq2_bs_sor_Nov18_rf, "plasma_ext_batch"))
fact[is.na(fact)] <- ""
sample_data(physeq2_bs_sor_Nov18_rf)$sample_type_batch <- as.factor(paste(get_variable(physeq2_bs_sor_Nov18_rf, "sample_type"), fact, sep = ""))

#Merge samples by newly created factor by calculating the mean of the OTU abundances
OTU_table_mean <- merge_samples(physeq2_bs_sor_Nov18_rf, "sample_type_batch", fun = mean)
otu_table(OTU_table_mean) <- t(otu_table(OTU_table_mean)) # Because merge_samples swaps the rows and columns of the otu table, transpose it back
# Generate a new metadata table that matches the newly created merged OTU table
metadata_phy_new <- sample_data(prune_samples(!duplicated(get_variable(physeq2_bs_sor_Nov18_rf, "sample_type_batch")), physeq2_bs_sor_Nov18_rf))
sample_names(metadata_phy_new) <- metadata_phy_new$sample_type_batch
sample_data(OTU_table_mean) <- metadata_phy_new
sample_names(OTU_table_mean) <- c("NTC","plasma-DENC batch A","plasma-DENC batch B","plasma batch A","plasma batch B","saliva","saliva-DENC","stool","stool-DENC")

## Transform the OTU-table to relative percentage
OTU_table_mean_perc <- transform_sample_counts(OTU_table_mean, function(OTU) 100 * OTU/sum(OTU))

#Tax glom at the Phylum level
OTU_table_mean_perc_phylum <- tax_glom(OTU_table_mean_perc, taxrank = "Phylum")
taxa_names(OTU_table_mean_perc_phylum) <- tax_table(OTU_table_mean_perc_phylum)[,"Phylum"]
# Pick top 20 Phyla
TopPhyla <- names(sort(taxa_sums(OTU_table_mean_perc_phylum), TRUE)[1:15])
OTU_table_mean_perc_phylum_top <- prune_taxa(TopPhyla, OTU_table_mean_perc_phylum)
# Create taxa-bar-plot
#sample_names(OTU_table_mean_perc_phylum_top) <- c("NTC","plasma-DENC batch A","plasma-DENC batch B","plasma batch A","plasma batch B","saliva","saliva-DENC","stool","stool-DENC")
x_axis_order <- c("stool","stool-DENC","saliva","saliva-DENC","plasma batch A","plasma-DENC batch A","plasma batch B","plasma-DENC batch B","NTC")
taxa_bar_plot <- plot_bar(OTU_table_mean_perc_phylum_top, fill = "Phylum") + scale_fill_manual(values=ARS_PerCol35) + scale_x_discrete(limits = x_axis_order) + ylab("Average abundance (%)") + xlab("Sample type")
taxa_plot_aes <- theme(axis.text.x = element_text(size=12, angle = 45, hjust=1, vjust=1), axis.text.y = element_text(size=12), legend.title=element_text(size=14, face="bold"), legend.text = element_text(size=14), axis.title = element_text(size=14, face="bold"))
taxa_bar_plot + taxa_plot_aes + ggtitle("Figure 3A. Taxonomic profiles of patient-matching plasma,\n stool and saliva samples and respective DENCs at the phylum level.")
ggsave("./taxonomy/Figure3A.pdf", width = 8, height = 6)

#Tax glom at the Class level
OTU_table_mean_perc_Class <- tax_glom(OTU_table_mean_perc, taxrank = "Class")
taxa_names(OTU_table_mean_perc_Class) <- tax_table(OTU_table_mean_perc_Class)[,"Class"]
# Pick top 20 Phyla
TopPhyla <- names(sort(taxa_sums(OTU_table_mean_perc_Class), TRUE)[1:15])
OTU_table_mean_perc_Class_top <- prune_taxa(TopPhyla, OTU_table_mean_perc_Class)
# Create taxa-bar-plot
taxa_bar_plot <- plot_bar(OTU_table_mean_perc_Class_top, fill = "Class")+ scale_fill_manual(values=ARS_PerCol35) + scale_x_discrete(limits = x_axis_order) + ylab("Average abundance (%)") + xlab("Sample type")
taxa_bar_plot + taxa_plot_aes + ggtitle("FigureS4A.Taxonomic profiles of patient-matching plasma,\n stool and saliva samples and respective DENCs")
ggsave("./taxonomy/FigureS4A.pdf", width = 8, height = 6)

#Tax glom at the Family level
OTU_table_mean_perc_Family <- tax_glom(OTU_table_mean_perc, taxrank = "Family")
#taxa_names(OTU_table_mean_perc_Family) <- tax_table(OTU_table_mean_perc_Family)[,"Family"]
# Pick top 20 Phyla
TopPhyla <- names(sort(taxa_sums(OTU_table_mean_perc_Family), TRUE)[1:20])
OTU_table_mean_perc_Family_top <- prune_taxa(TopPhyla, OTU_table_mean_perc_Family)
taxa_bar_plot <- plot_bar(OTU_table_mean_perc_Family_top, fill = "Family")+ scale_fill_manual(values=ARS_PerCol35) + scale_x_discrete(limits = x_axis_order) + ylab("Average abundance (%)") + xlab("Sample type")
taxa_bar_plot + taxa_plot_aes + ggtitle("FigureS4B.Taxonomic profiles of patient-matching plasma, stool and saliva samples and respective DENCs.")
ggsave("./taxonomy/FigureS4B.pdf", width = 8, height = 7)

#Tax glom at the Genus level
OTU_table_mean_perc_Genus <- tax_glom(OTU_table_mean_perc, taxrank = "Genus")
#taxa_names(OTU_table_mean_perc_Genus) <- tax_table(OTU_table_mean_perc_Genus)[,"Genus"]
# Pick top 20 Phyla
TopPhyla <- names(sort(taxa_sums(OTU_table_mean_perc_Genus), TRUE)[1:20])
OTU_table_mean_perc_Genus_top <- prune_taxa(TopPhyla, OTU_table_mean_perc_Genus)
taxa_bar_plot <- plot_bar(OTU_table_mean_perc_Genus_top, fill = "Genus")+ scale_fill_manual(values=ARS_PerCol35) + scale_x_discrete(limits = x_axis_order) + ylab("Average abundance (%)") + xlab("Sample type")
taxa_bar_plot + taxa_plot_aes + ggtitle("FigureS4C.Taxonomic profiles of patient-matching plasma, stool and saliva samples and respective DENCs.")
ggsave("./taxonomy/FigureS4C.pdf", width = 8, height = 7)
```

##Group 2

```{r echo=FALSE, results="hide", message=FALSE, fig.show="show",  warning=FALSE, fig.height=9, fig.width=8}
### JUL-19 DATA ###

#Subset OTU table to Jul-19 samples
physeq2_bs_sor_Jul19 <- subset_samples(physeq2_bs_sor, seq_run == "Jul_2019")
physeq2_bs_sor_Jul19 <- prune_taxa(taxa_sums(physeq2_bs_sor_Jul19) > 0, physeq2_bs_sor_Jul19)
# Sort samples by total read number:
as.data.frame(sort(sample_sums(physeq2_bs_sor_Jul19)))
# NORMALIZE OTUs by rarefaction to min. sample size
physeq2_bs_sor_Jul19_rf <- rarefy_even_depth(physeq2_bs_sor_Jul19, rngseed = 711, replace=FALSE, trimOTUs = TRUE, sample.size = 6042)

#Create new factor by combining plasma_ext_batch with sample_type
fact <- as.character(get_variable(physeq2_bs_sor_Jul19_rf, "plasma_ext_batch"))
fact[is.na(fact)] <- ""
sample_data(physeq2_bs_sor_Jul19_rf)$sample_type_batch <- as.factor(paste(get_variable(physeq2_bs_sor_Jul19_rf, "sample_type"), fact, sep = ""))

#Merge samples by newly created factor by calculating the mean of the OTU abundances
OTU_table_mean <- merge_samples(physeq2_bs_sor_Jul19_rf, "sample_type_batch", fun = mean)
otu_table(OTU_table_mean) <- t(otu_table(OTU_table_mean)) # Because merge_samples swaps the rows and columns of the otu table, transpose it back
# Generate a new metadata table that matches the newly created merged OTU table
metadata_phy_new <- sample_data(prune_samples(!duplicated(get_variable(physeq2_bs_sor_Jul19_rf, "sample_type_batch")), physeq2_bs_sor_Jul19_rf))
sample_names(metadata_phy_new) <- metadata_phy_new$sample_type_batch
sample_data(OTU_table_mean) <- metadata_phy_new
sample_names(OTU_table_mean) <- c("NTC","plasma-DENC batch C","plasma-DENC batch D","plasma-DENC batch E","plasma batch C","plasma batch D","plasma batch E")

## Transform the OTU-table to relative percentage
OTU_table_mean_perc <- transform_sample_counts(OTU_table_mean, function(OTU) 100 * OTU/sum(OTU))

#Tax glom at the Phylum level
OTU_table_mean_perc_phylum <- tax_glom(OTU_table_mean_perc, taxrank = "Phylum")
taxa_names(OTU_table_mean_perc_phylum) <- tax_table(OTU_table_mean_perc_phylum)[,"Phylum"]
# Pick top 20 Phyla
TopPhyla <- names(sort(taxa_sums(OTU_table_mean_perc_phylum), TRUE)[1:15])
OTU_table_mean_perc_phylum_top <- prune_taxa(TopPhyla, OTU_table_mean_perc_phylum)
# Create taxa-bar-plot
x_axis_order <- c("plasma batch C","plasma-DENC batch C","plasma batch D","plasma-DENC batch D","plasma batch E","plasma-DENC batch E","NTC") 
taxa_bar_plot <- plot_bar(OTU_table_mean_perc_phylum_top, fill = "Phylum") + scale_fill_manual(values=ARS_PerCol35) + scale_x_discrete(limits = x_axis_order) + ylab("Average abundance (%)") + xlab("Sample type")
taxa_bar_plot + taxa_plot_aes + ggtitle("FigureS6A.Taxonomic profile of plasma samples\n and their corresponding DENCs from DEBs C-E")
ggsave("./taxonomy/FigureS6.pdf", width = 8, height = 6)

#Tax glom at the Class level
OTU_table_mean_perc_Class <- tax_glom(OTU_table_mean_perc, taxrank = "Class")
taxa_names(OTU_table_mean_perc_Class) <- tax_table(OTU_table_mean_perc_Class)[,"Class"]
# Pick top 20 Phyla
TopPhyla <- names(sort(taxa_sums(OTU_table_mean_perc_Class), TRUE)[1:15])
OTU_table_mean_perc_Class_top <- prune_taxa(TopPhyla, OTU_table_mean_perc_Class)
# Create taxa-bar-plot
taxa_bar_plot <- plot_bar(OTU_table_mean_perc_Class_top, fill = "Class") + scale_fill_manual(values=ARS_PerCol35) + scale_x_discrete(limits = x_axis_order) + ylab("Average abundance (%)") + xlab("Sample type")
taxa_bar_plot + taxa_plot_aes + ggtitle("FigureS6B.Taxonomic profile of plasma samples\n and their corresponding DENCs from DEBs C-E")
ggsave("./taxonomy/FigureS6B.pdf", width = 8, height = 6)

#Tax glom at the Family level
OTU_table_mean_perc_Family <- tax_glom(OTU_table_mean_perc, taxrank = "Family")
#taxa_names(OTU_table_mean_perc_Family) <- tax_table(OTU_table_mean_perc_Family)[,"Family"]
# Pick top 20 Phyla
TopPhyla <- names(sort(taxa_sums(OTU_table_mean_perc_Family), TRUE)[1:20])
OTU_table_mean_perc_Family_top <- prune_taxa(TopPhyla, OTU_table_mean_perc_Family)
# Create taxa-bar-plot
taxa_bar_plot <- plot_bar(OTU_table_mean_perc_Family_top, fill = "Family") + scale_fill_manual(values=ARS_PerCol35) + scale_x_discrete(limits = x_axis_order) + ylab("Average abundance (%)") + xlab("Sample type")
taxa_bar_plot + taxa_plot_aes + ggtitle("FigureS6C.Taxonomic profile of plasma samples\n and their corresponding DENCs from DEBs C-E")
ggsave("./taxonomy/FigureS6C.pdf", width = 8, height = 7)

#Tax glom at the Genus level
OTU_table_mean_perc_Genus <- tax_glom(OTU_table_mean_perc, taxrank = "Genus")
#taxa_names(OTU_table_mean_perc_Genus) <- tax_table(OTU_table_mean_perc_Genus)[,"Genus"]
# Pick top 20 Phyla
TopPhyla <- names(sort(taxa_sums(OTU_table_mean_perc_Genus), TRUE)[1:20])
OTU_table_mean_perc_Genus_top <- prune_taxa(TopPhyla, OTU_table_mean_perc_Genus)
# Create taxa-bar-plot
taxa_bar_plot <- plot_bar(OTU_table_mean_perc_Genus_top, fill = "Genus") + scale_fill_manual(values=ARS_PerCol35) + scale_x_discrete(limits = x_axis_order) + ylab("Average abundance (%)") + xlab("Sample type")
taxa_bar_plot + taxa_plot_aes + ggtitle("FigureS6D.Taxonomic profile of plasma samples\n and their corresponding DENCs from DEBs C-E")
ggsave("./taxonomy/FigureS6D.pdf", width = 8, height = 7)
```

#Comparisons between sample-types

##Plasma_batch_A vs stool

```{r corr_plasmaA_vs_stool_var_dec, include=FALSE}
system("mkdir corr_sample_types")

# Subset OTU table to samples of batch A
physeq2_corr <- subset_samples(physeq2_bs_sor, plasma_ext_batch %in% c("A",NA) & sample_type %in% c("Plasma","Stool"))
physeq2_corr <- prune_taxa(taxa_sums(physeq2_corr) > 0, physeq2_corr)

# Normalize OTUs by rarefying at the minimum sample size.
physeq2_corr <- rarefy_even_depth(physeq2_corr, rngseed = 711, replace=FALSE, trimOTUs = TRUE) # Rarify OTU table to the minimum sample size.

sample_type1 <- "Plasma"
sample_type2 <- "Stool"

#contam_filters_results_column <- "plasmaA_vs_stool"
```

```{r corr_btwn_sample_types, echo=FALSE, message=FALSE, warning=FALSE, results="show"}
# Get the metadata of sample_type1
metadata_corr_sample_type1 <- subset(sample_data(physeq2_corr), sample_type == sample_type1)
# Get the metadata of sample_type2
metadata_corr_sample_type2 <- subset(sample_data(physeq2_corr), sample_type == sample_type2)

# Get the "matching samples ID" of the sample_type1 samples
matching_samples_sample_type1 <- levels(metadata_corr_sample_type1$matching_samples)
# Get the "matching samples ID" of the sample_type2 samples
matching_samples_sample_type2 <- levels(metadata_corr_sample_type2$matching_samples)
# Get the "matching samples ID" shared between these two sets of sample_type1 and sample_type2 samples
matching_samples_sample_type1_sample_type2 <- intersect(matching_samples_sample_type1, matching_samples_sample_type2)

#Subset phyloseq object to matching sample_type1 and sample_type2 samples
physeq2_corr_MS <- prune_samples(get_variable(physeq2_corr, "matching_samples") %in% matching_samples_sample_type1_sample_type2, physeq2_corr)

# Extract OTU table of matching sample_type1 samples
OTU_table_corr_st1 <- subset_samples(physeq2_corr_MS, sample_type == sample_type1)
OTU_table_corr_st1.df <- otu_table(OTU_table_corr_st1)
OTU_table_corr_st1.df.sort <- OTU_table_corr_st1.df[,order(get_variable(OTU_table_corr_st1, "matching_samples"))]
OTU_table_corr_st1.df.sort.PA <- transform_sample_counts(OTU_table_corr_st1.df.sort, function(abund) 1*(abund>0)) # Transform extracted OTU table into presence/absence

# Extract OTU table of matching sample_type2 samples
OTU_table_corr_st2 <- subset_samples(physeq2_corr_MS, sample_type == sample_type2)
OTU_table_corr_st2.df <- otu_table(OTU_table_corr_st2)
OTU_table_corr_st2.df.sort <- OTU_table_corr_st2.df[,order(get_variable(OTU_table_corr_st2, "matching_samples"))]
OTU_table_corr_st2.df.sort.PA <- transform_sample_counts(OTU_table_corr_st2.df.sort, function(abund) 1*(abund>0)) # Transform extracted OTU table into presence/absence

# Get the names of the OTUs shared between sample-types per sample
all_shared_OTUs_names <- list()
for(i in 1:ncol(OTU_table_corr_st1.df.sort.PA)){
  x <- OTU_table_corr_st1.df.sort.PA[,i] * OTU_table_corr_st2.df.sort.PA[,i]
  all_shared_OTUs_names[[colnames(OTU_table_corr_st1.df.sort.PA[,i])]] <- as.character(rownames(x)[x==1])
}

# Get the OTU abundances in each "pair of compared sample-types" per sample
all_shared_OTUs_abund <- all_shared_OTUs_names
for(i in 1:length(all_shared_OTUs_names)){
  if(length(all_shared_OTUs_names[[i]]) > 0){
      sample <- colnames(OTU_table_corr_st1.df.sort[,i])
      all_shared_OTUs_abund[[sample]] <- cbind(tax_table(physeq2_corr_MS)[all_shared_OTUs_names[[i]],"Genus"], as.data.frame(OTU_table_corr_st1.df.sort[all_shared_OTUs_names[[i]], i]), as.data.frame(OTU_table_corr_st2.df.sort[all_shared_OTUs_names[[i]], i]))
  }
}

# Get the total number of shared OTUs per sample 
all_shared_OTU_nums <- sapply(all_shared_OTUs_names, length)
# Get sample_type 1 richness
st1_richness <- estimate_richness(OTU_table_corr_st1, measures = "Observed")
# Get the total number of reads of the "shared OTUs" per sample. NOTE: plasma has to be sample_type 1.
all_shared_OTUs_reads <- sapply(all_shared_OTUs_abund, function(x){
  if(length(x) > 0){
    sum(x[,2])
  }else{
    length(x)  
  }
})
# Get the total number of reads per sample
st1_reads_per_sample <- sample_sums(OTU_table_corr_st1)

# Create table with total number of sample_type_1 (plasma) OTUs, percentage of OTUs and percentage of reads
all_shared_OTUs_table <- cbind(all_shared_OTU_nums,  100*all_shared_OTU_nums/st1_richness, 100*all_shared_OTUs_reads/st1_reads_per_sample)
names(all_shared_OTUs_table) <- c("num_of_OTUs","perc_of_OTUs","perc_of_reads")
all_shared_OTUs_table <- round(all_shared_OTUs_table, digits=1)
kable(all_shared_OTUs_table, caption = "Total number of plasma OTUs shared between matching samples")
```

```{r include=FALSE}
all_shared_OTUs_table.plA_vs_st <- all_shared_OTUs_table
all_shared_OTUs_names.plA_vs_st <- all_shared_OTUs_names
all_shared_OTUs_abund.plA_vs_st <- all_shared_OTUs_abund
plA_mtch_samples_richness <- st1_richness
plA_mtch_samples_reads <- st1_reads_per_sample
```

##Plasma_batch_B vs stool correlation

```{r corr_plasmaB_vs_stool_var_dec, include=FALSE}
# Subset OTU table to samples of batch A
physeq2_corr <- subset_samples(physeq2_bs_sor, plasma_ext_batch %in% c("B",NA) & sample_type %in% c("Plasma","Stool"))
physeq2_corr <- prune_taxa(taxa_sums(physeq2_corr) > 0, physeq2_corr)

# Normalize OTUs by rarefying at the minimum sample size.
physeq2_corr <- rarefy_even_depth(physeq2_corr, rngseed = 711, replace=FALSE, trimOTUs = TRUE) # Rarify OTU table to the minimum sample size.

sample_type1 <- "Plasma"
sample_type2 <- "Stool"

#contam_filters_results_column <- "plasmaB_vs_stool"
```

```{r ref.label='corr_btwn_sample_types', echo=FALSE}
```

```{r include=FALSE}
all_shared_OTUs_table.plB_vs_st <- all_shared_OTUs_table
all_shared_OTUs_names.plB_vs_st <- all_shared_OTUs_names
all_shared_OTUs_abund.plB_vs_st <- all_shared_OTUs_abund
plB_mtch_samples_richness <- st1_richness
plB_mtch_samples_reads <- st1_reads_per_sample
```

##Plasma_batch_A vs saliva correlation

```{r corr_plasmaA_vs_saliva_var_dec, include=FALSE}
# Subset OTU table to samples of batch A
physeq2_corr <- subset_samples(physeq2_bs_sor, plasma_ext_batch %in% c("A",NA) & sample_type %in% c("Plasma","Saliva"))
physeq2_corr <- prune_taxa(taxa_sums(physeq2_corr) > 0, physeq2_corr)

# Normalize OTUs by rarefying at the minimum sample size.
physeq2_corr <- rarefy_even_depth(physeq2_corr, rngseed = 711, replace=FALSE, trimOTUs = TRUE) # Rarify OTU table to the minimum sample size.

sample_type1 <- "Plasma"
sample_type2 <- "Saliva"

#contam_filters_results_column <- "plasmaA_vs_saliva"
```

```{r ref.label='corr_btwn_sample_types', echo=FALSE}
```

```{r include=FALSE}
all_shared_OTUs_table.plA_vs_sl <- all_shared_OTUs_table
all_shared_OTUs_names.plA_vs_sl <- all_shared_OTUs_names
all_shared_OTUs_abund.plA_vs_sl <- all_shared_OTUs_abund
plA_mtch_samples_richness <- st1_richness
plA_mtch_samples_reads <- st1_reads_per_sample
```

##Plasma_batch_B vs saliva correlation

```{r corr_plasmaB_vs_saliva_var_dec, include=FALSE}
# Subset OTU table to samples of batch B
physeq2_corr <- subset_samples(physeq2_bs_sor, plasma_ext_batch %in% c("B",NA) & sample_type %in% c("Plasma","Saliva"))
physeq2_corr <- prune_taxa(taxa_sums(physeq2_corr) > 0, physeq2_corr)

# Normalize OTUs by rarefying at the minimum sample size.
physeq2_corr <- rarefy_even_depth(physeq2_corr, rngseed = 711, replace=FALSE, trimOTUs = TRUE) # Rarify OTU table to the minimum sample size.

sample_type1 <- "Plasma"
sample_type2 <- "Saliva"

corr_plot_file <- "./corr_sample_types/plasmaB_vs_saliva_corr_Rho0.2.pdf"
contam_filters_results_column <- "plasmaB_vs_saliva"
```

```{r ref.label='corr_btwn_sample_types', echo=FALSE}
```

```{r include=FALSE}
all_shared_OTUs_table.plB_vs_sl <- all_shared_OTUs_table
all_shared_OTUs_names.plB_vs_sl <- all_shared_OTUs_names
all_shared_OTUs_abund.plB_vs_sl <- all_shared_OTUs_abund
plB_mtch_samples_richness <- st1_richness
plB_mtch_samples_reads <- st1_reads_per_sample
```

##DEB_A and DEB_B average

```{r Average_OTUs_shared_pl_st, echo=FALSE}
# Edits of all_shared_OTUs_table.plA_vs_st
# Add the table's rownames as a column and add the "matching samples" IDs as table-names
all_shared_OTUs_table.plA_vs_st <- cbind(biological_sample = rownames(all_shared_OTUs_table.plA_vs_st), all_shared_OTUs_table.plA_vs_st)
matching_samples <- sample_data(physeq2_bs_sor)[row.names(all_shared_OTUs_table.plA_vs_st),"matching_samples"]
matching_samples <- as.character(matching_samples$matching_samples)
rownames(all_shared_OTUs_table.plA_vs_st) <- matching_samples

# Edits of all_shared_OTUs_table.plB_vs_st
# Add the table's rownames as a column and add the "matching samples" IDs as table-names
all_shared_OTUs_table.plB_vs_st <- cbind(biological_sample = rownames(all_shared_OTUs_table.plB_vs_st), all_shared_OTUs_table.plB_vs_st)
matching_samples <- sample_data(physeq2_bs_sor)[row.names(all_shared_OTUs_table.plB_vs_st),"matching_samples"]
matching_samples <- as.character(matching_samples$matching_samples)
rownames(all_shared_OTUs_table.plB_vs_st) <- matching_samples

# Create a version of all_shared_OTUs_table.plB_vs_st in which the samples present in DEB A but not in B, get inserted. This for the later calculation of the average of the columns of all_shared_OTUs_table.plA_vs_st and all_shared_OTUs_table.plB_vs_st. In this way, when a samples is only present in DEB A, the average with result in the original DEB A value.
all_shared_OTUs_table.plB_vs_st.m <- all_shared_OTUs_table.plA_vs_st
all_shared_OTUs_table.plB_vs_st.m[rownames(all_shared_OTUs_table.plB_vs_st),2:4] <- all_shared_OTUs_table.plB_vs_st[,2:4]

# Table with the average number of OTUs shared between plasma and stool in DEB A and B
all_sh_OTUs_tbl.plA_B_vs_st <- cbind(rowMeans(cbind(all_shared_OTUs_table.plA_vs_st$num_of_OTUs, all_shared_OTUs_table.plB_vs_st.m$num_of_OTUs)), rowMeans(cbind(all_shared_OTUs_table.plA_vs_st$perc_of_OTUs, all_shared_OTUs_table.plB_vs_st.m$perc_of_OTUs)), rowMeans(cbind(all_shared_OTUs_table.plA_vs_st$perc_of_reads, all_shared_OTUs_table.plB_vs_st.m$perc_of_reads)))
colnames(all_sh_OTUs_tbl.plA_B_vs_st) <- colnames(all_shared_OTUs_table.plA_vs_st)[-1]
rownames(all_sh_OTUs_tbl.plA_B_vs_st) <- rownames(all_shared_OTUs_table.plA_vs_st)

kable(all_sh_OTUs_tbl.plA_B_vs_st, caption = "Average number of OTUs and reads shared between plasma and stool in DEB A and B. Info for Table S2.")
```

```{r Average_OTUs_shared_pl_sl, echo=FALSE}
# Edits of all_shared_OTUs_table.plA_vs_sl
# Add the table's rownames as a column and add the "matching samples" IDs as table-names
all_shared_OTUs_table.plA_vs_sl <- cbind(biological_sample = rownames(all_shared_OTUs_table.plA_vs_sl), all_shared_OTUs_table.plA_vs_sl)
matching_samples <- sample_data(physeq2_bs_sor)[row.names(all_shared_OTUs_table.plA_vs_sl),"matching_samples"]
matching_samples <- as.character(matching_samples$matching_samples)
rownames(all_shared_OTUs_table.plA_vs_sl) <- matching_samples

# Edits of all_shared_OTUs_table.plB_vs_sl
# Add the table's rownames as a column and add the "matching samples" IDs as table-names
all_shared_OTUs_table.plB_vs_sl <- cbind(biological_sample = rownames(all_shared_OTUs_table.plB_vs_sl), all_shared_OTUs_table.plB_vs_sl)
matching_samples <- sample_data(physeq2_bs_sor)[row.names(all_shared_OTUs_table.plB_vs_sl),"matching_samples"]
matching_samples <- as.character(matching_samples$matching_samples)
rownames(all_shared_OTUs_table.plB_vs_sl) <- matching_samples

# Create a version of all_shared_OTUs_table.plB_vs_sl in which the samples present in DEB A but not in B, get inserted. This for the later calculation of the average of the columns of all_shared_OTUs_table.plA_vs_sl and all_shared_OTUs_table.plB_vs_sl. In this way, when a samples is only present in DEB A, the average with result in the original DEB A value.
all_shared_OTUs_table.plB_vs_sl.m <- all_shared_OTUs_table.plA_vs_sl
all_shared_OTUs_table.plB_vs_sl.m[rownames(all_shared_OTUs_table.plB_vs_sl),2:4] <- all_shared_OTUs_table.plB_vs_sl[,2:4]

# Table with the average number of OTUs shared between plasma and saliva in DEB A and B
all_sh_OTUs_tbl.plA_B_vs_sl <- cbind(rowMeans(cbind(all_shared_OTUs_table.plA_vs_sl$num_of_OTUs, all_shared_OTUs_table.plB_vs_sl.m$num_of_OTUs)), rowMeans(cbind(all_shared_OTUs_table.plA_vs_sl$perc_of_OTUs, all_shared_OTUs_table.plB_vs_sl.m$perc_of_OTUs)), rowMeans(cbind(all_shared_OTUs_table.plA_vs_sl$perc_of_reads, all_shared_OTUs_table.plB_vs_sl.m$perc_of_reads)))
colnames(all_sh_OTUs_tbl.plA_B_vs_sl) <- colnames(all_shared_OTUs_table.plA_vs_sl)[-1]
rownames(all_sh_OTUs_tbl.plA_B_vs_sl) <- rownames(all_shared_OTUs_table.plA_vs_sl)

kable(all_sh_OTUs_tbl.plA_B_vs_sl, caption = "Average number of OTUs and reads shared between plasma and saliva in DEB A and B. Info for Table S2.")
```

```{r include=FALSE}
save.image("/Users/zozaya-valdesenrique/Documents/work/bioinformatics/experiments/16S_rRNA_gene/GB_revision1/R scripts for Github/Group 1 and 2/div_analysis/Group_1_and_2.diversity_analysis.RData")
save(physeq2_bs, physeq2_bs_sor, physeq2_bs_sor_pl_bl, file = "/Users/zozaya-valdesenrique/Documents/work/bioinformatics/experiments/16S_rRNA_gene/GB_revision1/R scripts for Github/Group 1 and 2/div_analysis/Group_1_and_2.OTU_tables.RData")
```
